\bf Model Visualization Techniques for a Social Network Model

### 2017-06-05 paper.Rnw
done with the edits from last week

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators. Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
}

\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore the fitting process rather than just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out of the black box and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks for these estimates rely on simulations from the fitted model. Again, these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring these elements and others into the foreground, leading to a better understanding and higher accessibility of stochastic actor-oriented models  for social network analysts.

@@

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \cite{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five edges:  $\{2 \mapto 4, 3 \mapto 4, 1 \mapto 5, 3 \mapto 5, 5 \mapto 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

@@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. But because there is no natural placement of the points, a random placement is used, then adjusted iteratively via a layout algorithm, of which there are many. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.

Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. The network's edges act as springs pushing and pulling the nodes in a lower dimensional (usually two-dimensional) space. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes. \hh{XXX how does the adjacency matrix feature here? The adjacency matrix is  a different mathematical representation of a network, but describes the same network as a node-link diagram.} 
Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems. Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix.

The second primary method for network visualization uses the adjacency matrix of the network. \hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} An example of this type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example. This method is particularly useful when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs. \hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our undirected toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id, shape =group, color = group), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .02) +
  scale_color_manual(values = c("red", "blue")) + 
  theme_net() + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = 0.5, yend=5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=y, yend=yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = 0:5+.5, yend= 0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1) +
  xlab("from (ego)") +
  ylab("to (alter)")
@
 \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process we usually condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), which we will refer to as $X(T)$. This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$. All other parts of the process are unseen. We model the process $X(T)$ as a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie. Once an actor is given the chance to change a tie, it tries to maximize a utility function based on the current and potential future states of the network.

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\alpha$. For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$. \hh{XXX that means that the function is essentially just a rate parameter, right?}
Other modelling scenarios allow this rate to be more flexible,  e.g.\  be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
\hh{How is the rate parameter different from a rate parameter?}
The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$. We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$. This assumption leads to the rate for the whole network, $n\alpha$. In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time. The objective function is then written as \hh{the sum of the individual changes XXX write out the math in words XXX} $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p. 371]{snijders01}. \hh{'There are many' another vague phrase} There are many possible parameters $\beta_i$ to add to the model. \hh{'They' ... be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network. \sct{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$. The second set of effects are the actor-level or covariate effects.} \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} These effects also depend on the structure of the network. They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can be included \hh{currently} in the objective function, see \citet{RSiena}.

>>>>

\st{
%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
Unlike other network models, SAOMs are not very well understood. Traditional modeling steps, like determining the likelihood function, are very complex. Because of this, SAOMs are fit to data using a series of MCMC phases for finding method of moments estimators. Rather than treat SAOM fits and the fitting process as black boxes, we combine the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures ``behind the scenes" of SAOMs, we help researchers working with the models better understand the implications and analysis of these models.
}

\par The guiding principles of model visualization introduced in \citet{modelvis} are (1) display the model in the data space, (2) collections of models are more informative than singletons, and (3) explore how algorithms work, instead of just looking at the final result. Stochastic actor-oriented models are a prime example of a set of models that  benefit greatly from application of model visualization. First, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden in the model fitting process. Bringing the CTMC out of the black box and into the light of visualization  provides researchers with insights into the underlying features of the model. Secondly, SAOMs allow for a great deal of parameters that can be added to the objective function, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, and so are the associated parameters. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects. Finally, estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and convergence checks on these estimates rely on simulations from the fitted model. Again, these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring these elements and others into the foreground, leading to a better understanding and higher accessibility of stochastic actor-oriented models  for social network analysts.

@@

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

@@

The first method, the node-link diagram, represents nodes with markers in 2D Euclidean space and then represents edges by connecting the markers with lines when there is an edge between the two nodes. But because there is no natural placement of the points, a random placement is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

Some layout algorithms were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. The network's edges act as springs pushing and pulling the nodes in a lower dimensional (usually two-dimensional) space. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes. \hh{XXX how does the adjacency matrix feature here? The adjacency matrix is  a different mathematical representation of a network, but describes the same network as a node-link diagram.} 
Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, treat networks like physical systems. Other algorithms are based on multi-dimensional scaling, which relies on a defined distance matrix, or an eigen value, which depends on the eigenvalues of the network's adjacency matrix \citep{drawingalgs}. The node-link diagram for our toy network is shown in Figure~\ref{fig:toyplots}.

The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as 
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists} i\to j \\
                                   0 & otherwise
  \end{cases}
\]
In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.
%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} 
This type of visualization is also shown in Figure~\ref{fig:toyplots} for our toy example.  %\hh{XXX What is the exact logic here that Ghoniem used to come to the conclusion that edges in undirected graphs are over-represented? Why would directed graphs make that better? wouldn't that just add to the number of edges? } This, however, may actually be an advantage for \emph{directed} graphs, where exactly the correct number of edges is represented in a matrix visualization, but a node-link diagram may underrepresent the edge count with overlapping edges.

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\linewidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) + 
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 10, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, arrowgap = .05, directed = T) +
  theme_net() + 
  labs(title = "Node-Link Diagram")

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = -to), color = 'grey20', fill = 'black') +
  #geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = -0.5, yend=-5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=-y, yend=-yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = -0:5+.5, yend= -0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5), position = "top") +
  scale_y_continuous(breaks = (-5):(-1), labels = 5:1, limits = c(-5.5,-.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization", y = "to") + 
  theme(aspect.ratio = 1) 
@
% \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

Each type of visualization comes with its own advantages and disadvantages. Paths between nodes are easier to determine with node-link diagrams than with adjacency matrix visualizations. In node-link diagrams, node information can be incorporated into the visualization by coloring or changing the shape of the points representing the nodes. This is not necessarily possible in matrix representation, which is more focused on edges. Adjacency matrix visualization is particularly useful, however, when the network is very complex, dense, or large. Experimental studies have shown this to be true. For example, with seven basic perceptual tasks on networks, including things like node and edge count, adjacency matrix visualizations outperform node-link diagrams as the size and density of the network increases \citep{adjmatviz}. One drawback of the adjacency matrix visualization that \citeauthor{adjmatviz} found was that edges are overrepresented for undirected graphs, due to the symmetry of $\mathbf{A}$. There is not one correct way to visualize network information, and we will be using both of the methods throughout this paper.

@@

In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$. We denote these network observations $x(t_1), \dots, x(t_M)$, and in modelling we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), call it $X(T)$. This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$. All other parts of the process are unseen. The process $X(T)$ is a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie. Once an actor is given the chance to change a tie, it tries to maximize a sort of utility function based on the current and potential future states of the network.

@@

For the network $x$ and each actor $i$ in the network, the rate function dictates how often the actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network. For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$. In other modelling scenarios, this rate can be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$. We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$. This assumption leads to the rate for the whole network, $n\alpha$. In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$

@@

Thanks to the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time. The objective function is written as $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p. 371]{snijders01}. There are many possible parameters $\beta_i$ to add to the model. They can be split up into two groups: first, the structural effects, which only depend on the structure of the network. The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$. The second set of effects are the actor-level or covariate effects. These effects also depend on the structure of the network. They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can be included in the objective function, see \citet{RSiena}.
